Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
3-steps finito

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting.\todo{[A] direi: ``...
    in the context of the Double Pushout (DPO) Approach to rewriting in adhesive categories.''}
    More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of distributed systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently from more recent approaches
like, e.g., SqPO~\cite{CorradiniHHK06}
or PBPO$^+$~\cite{OverbeekER23}) and to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by what is called the
Church-Rosser theorem of DPO rewriting~\cite[Section~4]{Ehrig78}.

@@

So far, most of the theoretical results focussed on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagrams mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expressions, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, one is required to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering general
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{BehrHK23}.

@@

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{Handbook, BaldanCHKS06, tobias}.\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weaker form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the ``Double-Pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

@@

Intuitively, the existence of the independence pair
captures the possibility of switching the application of the two rules:
$f_0 \circ i_1$ is a match for $\rho_1$ in $G_0$ and the
existence of $i_0 : R_0 \to D_1$ means that $\rho_1$ does not delete
items in the image of $K_0$, thus $\rho_0$ can be applied after
$\rho_1$.

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching'', i.e., which differs
only for the order of indepedent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ \todo{A: ma manca ``and  $\dder{D}_j = \dder{E}_j$ for each $j \not \in \{i,i+1\}$''??} we write
  $\der{D} \shift{\transp{i}} \der{E}$.

  A \emph{switching sequence} is  a sequence $\{\der{D}_{k}\}_{k=0}^m$ of derivations such that 
  \[\der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
  \der{D}_m\]
 In such case, we will\todo{A: io cambierei tutti i ``we will'' in ``we''.} often denote by $\nu_{k}$ the transposition $\transp{i_k}$. Given $k \leq h$,
 we denote by $\nu_{k,h}$ the composition
 $\nu_h \circ \nu_{h-1} \circ \ldots \nu_k$. We will say that the
 switching sequence consists of inversions if for all
 $k \in \interval[0]{m}$ the transposition $\nu_k$ is an inversion for
 $\nu_{1,m}$.

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ consists of three rules as
  depicted in Fig.~\ref{fi:rules}, where numbers are used to represent
  the mappings from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node, while $\rho_1$ deletes an edge. Instead, $\rho_2$ is not
  linear as it ``merges'' two nodes.

@@

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a non-strong enforcing
  adhesive rewriting system. Instead every left-linear system in category $\cat{Graph}$ is
  strong enforcing, by the results of Appendix~\ref{app:fill}, summarized in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

		We have to construct the two dotted arrows in the diagram below.
		\[\xymatrix@C=15pt{L_0 \ar[d]_{z'_0\circ o_0}&& K_0 \ar[d]_{n_1\circ w}\ar[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@{.>}@/^.35cm/[drrr]_(.4){\alpha_1}|(.285)\hole \ar[dr]|(.28)\hole_{y_1\circ e_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){\alpha_2} \ar[dl]|(.28)\hole^{y_1\circ b_1}& K_1 \ar[d]^{d_2\circ u}\ar[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{g_2\circ a_1}\\G'_1 && \ar[ll]^{z_1} Q_5 \ar[rr]_{z'_1}&& G'_2  && \ar[ll]^{x_2} Q_3 \ar[rr]_{y_2}&& G_2}\]

@@

The following proposition establishes a key property of $\mathcal{A}$-Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks \cite{ehrig2004adhesive,BehrHK23,lack2005adhesive}.

@@

\begin{definition} Let $\X$ be an $\mathcal{M}$-adhesive category, we say that the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}^+$} if (at least) one of the following holds true:
	\begin{enumerate}
		\item $(\X, \mathcal{M})$ is of type $\mathbb{B}$;
	\item $\X$ is $\Y/Y$ and $\mathcal{M}=\mathcal{N}/Y$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$;
 	\item $\X$ is $Y/\Y$ and $\mathcal{M}=Y/\mathcal{N}$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$:
 	\item $\X$ is $\Y^{\A}$ and $\mathcal{M}=\mathcal{N}^{\mathcal{A}}$ for some category category $\A$ and $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$.
	\end{enumerate}
\end{definition}
\begin{example}
  \label{ex:graph-type-bb}
	The pair $(\mathbf{Graph}, \mon(\mathbf{Graph}))$ made by the topos of graphs and its class of mons is of type $\mathbb{B}^+$  but not of type $\mathbb{B}$. Indeed not every monomorphism of graphs is a coproduct coprojection.
\end{example}

@@

Using \Cref{thm:good1} we finally get the main result of our appendix.
\begin{corollary}
  \label{cor:llr-is-strEnf}
Let $(\X, \mathcal{M})$ be a pair of type $\mathbb{B}^+$, then every left-linear DPO-rewriting system is strong enforcing.
\end{corollary}

@@

\begin{lemma}\label{lem:idevero}
	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and $\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ be two non-empty derivations in a linear DPO-rewriting system $(\X, \R)$ . Let  $\sigma:[0,n-1]\to [0,n-1]$  is a consistent permutation and suppose that the set 
	\[E_{\sigma}=\{i\in [0,n-1]\mid \sigma(i)=i\}\]
	is not empty with maximum $s_\sigma$. Then $\{\dder{D}_i\}_{i=0}^{s_\sigma}$ and $\{\der{D}'_i\}_{i=0}^{s_\sigma}$ are abstraction equivalent.  
\end{lemma}

@@

	\begin{enumerate}
		\item We have to construct the two dotted arrows in the diagram
		below.
		\[\xymatrix@C=15pt{L_0 \ar[d]_{z'_0\circ o_0}&& K_0
			\ar[d]_{n_1\circ w_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
			\ar@{.>}@/^.35cm/[drrr]_(.4){\beta_0}|(.285)\hole
			\ar[dr]|(.28)\hole_{y_1\circ e_0} && L_1
			\ar@{.>}@/_.35cm/[dlll]^(.4){\beta_1}
			\ar[dl]|(.28)\hole^{y_1\circ b_0}& K_1 \ar[d]^{d_1\circ
				u_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{g_2\circ
				a_1}\\G'_1 && \ar@{>->}[ll]^{z_1} Q_5 \ar[rr]_{z'_1}&& G'_2 &&
			\ar@{>->}[ll]^{x_2} Q_3 \ar[rr]_{y_2}&& G_3}\]

@@

		Our aim is to construct the dotted arrow in the following diagram.
		\[\xymatrix@C=15pt{L_1 \ar[d]_{f_0\circ i_0}&& K_1
			\ar[d]_{q_0\circ v_1}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_1
			\ar@{.>}@/^.35cm/[drrr]_(.4){\epsilon_0}|(.285)\hole
			\ar[dr]|(.28)\hole_{s_1\circ j_1} && L_2
			\ar@{.>}@/_.35cm/[dlll]^(.4){\epsilon_1}
			\ar[dl]|(.28)\hole^{s_1\circ \alpha_1}& K_2
			\ar[d]^{\lambda_0\circ \phi_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} &&
			R_2 \ar[d]^{\xi_1\circ \gamma_1}\\G_0 && \ar@{>->}[ll]^{s_0} Q_0
			\ar[rr]_{t_0}&& H'_1 && \ar@{>->}[ll]^{\xi_0} Q_6
			\ar[rr]_{\xi'_0}&& H'_2}\]

@@

\begin{proof}
	To fix the notation, let us depict the derivation $\der{D}=\{\dder{D}_i\}_{i=0}^2$:
	\[\xymatrix@C=13pt{L_0\ar[d]_{m_0}&& K_0 \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@/^.35cm/[drrr]_(.4){i_0}|(.285)\hole \ar[dr]|(.28)\hole_{h_0} && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.28)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[r]^{r_1} & R_1 \ar[dr]|(.28)\hole_{h_1} \ar@/^.35cm/[drrr]_(.4){j_0}|(.285)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){j_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]_{h_2} \\G_0 && \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1  && \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2 && \ar@{>->}[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]

@@

	We begin switching $\dder{D}_0$ and $\dder{D}_1$ to get the following diagram.
	\[\xymatrix@C=15pt{L_1\ar[d]^{f_0\circ i_1}&& K_1 \ar[d]_{k'_1}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){a_0}|(.285)\hole \ar[dr]|(.28)\hole_{h'_1} && L_0 \ar@/_.35cm/[dlll]^(.4){a_1} \ar[dl]|(.28)\hole^{m'_0}& K_0 \ar[d]^{k'_0}\ar@{>->}[l]_{l_0} \ar[r]^{r_0} & R_0 \ar[dr]|(.28)\hole_{g_1\circ i_0} \ar@/^.35cm/[drrr]_(.4){b_0}|(.285)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){b_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]_{h_2} \\G_0 && \ar@{>->}[ll]^{f'_0} D'_0 \ar[rr]_{g'_0}&& G'_1  && \ar@{>->}[ll]^{f'_1} D'_1 \ar[rr]_{g'_1}&& G_2 && \ar@{>->}[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]

>>>>

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of distributed systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{CMREHL:AAGT} 
is nowadays considered the standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently e.g. from the SqPO 
approach~\cite{CorradiniHHK06}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by what is called the
Church-Rosser theorem of DPO rewriting~\cite[Section~xxx]{CMREHL:AAGT}.

@@

So far, most of the theoretical results focussed on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagram mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering general
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{BehrHK23}.

@@

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the a characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{Handbook, tobias}.\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the ``double pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

@@

Intuitively, the existence of the independence pair
captures the possibility of switching the application of the two rules:
$f_0 \circ i_1$ is a match for $\rho_1$ in $G_0$ and the
existence of $i_0 : R_0 \to D_1$ indicates that $\rho_1$ does not delete
items in the image of $K_0$ and thus $\rho_0$ can be applied after
$\rho_1$.

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching'', i.e., which differs
only for the order of indepedent steps. The same notion can be given
for left-linear rewriting systems by relying on notion of switch.

@@

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ we write
  $\der{D} \shift{\transp{i}} \der{E}$.

  A \emph{switching sequence} is  a sequence $\{\der{D}_{k}\}_{i=0}^m$ of derivations such that 
  \[\der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
  \der{D}_m\]
 In such case, we will often denote by $\nu_{k}$ the transposition $\transp{i_k}$. Given $k \leq h$,
 we denote by $\nu_{k,h}$ the composition
 $\nu_h \circ \nu_{h-1} \circ \ldots \nu_k$. We will say that the
 switching sequence consists of inversions if for all
 $k \in \interval[0]{m}$ the transposition $\nu_k$ is an inversion for
 $\nu_{1,m}$.

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ consists of three rules as
  depicted in Fig.~\ref{fi:rules}, where numbers are used to represent
  the mappings from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node, while $\rho_1$ deletes an edge. Instead, $\rho_2$ is non
  linear as it ``merges'' two nodes.

@@

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a non strong enforcing
  adhesive rewriting system. Instead the category $\cat{Graph}$ is
  strong enforcing by the result in .... 
\end{example}

@@

		We have to construct the two dotted arrows in the diagram below.
		\[\xymatrix@C=15pt{L_0 \ar[d]_{z'_0\circ o_0}&& K_0 \ar[d]_{n_1\circ w}\ar[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@{.>}@/^.35cm/[drrr]_(.4){\alpha_1}|(.29)\hole \ar[dr]|(.28)\hole_{y_1\circ e_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){\alpha_2} \ar[dl]|(.28)\hole^{y_1\circ b_1}& K_1 \ar[d]^{d_2\circ u}\ar[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{g_2\circ a_1}\\G'_1 && \ar[ll]^{z_1} Q_5 \ar[rr]_{z'_1}&& G'_2  && \ar[ll]^{x_2} Q_3 \ar[rr]_{y_2}&& G_2}\]

@@

The following proposition establishes a key property of $\mathcal{A}$-Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks \cite{ehrig2004adhesive,behr2022fundamentals,lack2005adhesive}.

@@

\begin{definition} Let $\X$ be an $\mathcal{M}$-adhesive category, we say that the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}^+$} if (at least) one of the following holds true:
	\begin{enumerate}
		\item $(\X, \mathcal{M})$ is of type $\mathbb{B}$;
	\item $\X$ is $\Y/Y$ and $\mathcal{M}=\mathcal{N}/Y$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$;
 	\item $\X$ is $Y/\Y$ and $\mathcal{M}=Y/\mathcal{N}$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$:
 	\item $\X$ is $\Y^{\A}$ and $\mathcal{M}=\mathcal{N}^{\mathcal{A}}$ for some category category $\A$ and $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$.
	\end{enumerate}
\end{definition}
\begin{example}
	The pair $(\mathbf{Graph}, \mon(\mathbf{Graph}))$ made by the topos of graphs and its class of mons is of type $\mathbb{B}^+$  but not of type $\mathbb{B}$. Indeed not every monomorphism of graphs is a coproduct coprojection.
\end{example}

@@

Using \Cref{thm:good1} we finally get the main result of our appendix.
\begin{corollary}
Let $(\X, \mathcal{M})$ be a pair of type $\mathbb{B}^+$, then every left-linear DPO-rewriting system is strong enforcing.
\end{corollary}

@@

\begin{lemma}\label{lem:idevero}
	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and $\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ be two non-empty derivations in a linear DPO-rewriting system $(\X, \R)$ . Let  $\sigma:[0,n-1]\to [0,n-1]$  is a consistent permutation and suppose that the set 
	\[E_{\sigma}=\{i\in [0,n-1]\mid \sigma(i)=i\}\]
	is non empty with maximum $s_\sigma$. Then $\{\dder{D}_i\}_{i=0}^{s_\sigma}$ and $\{\der{D}'_i\}_{i=0}^{s_\sigma}$ are abstraction equivalent.  
\end{lemma}

@@

	\begin{enumerate}
		\item We have to construct the two dotted arrows in the diagram
		below.
		\[\xymatrix@C=15pt{L_0 \ar[d]_{z'_0\circ o_0}&& K_0
			\ar[d]_{n_1\circ w_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
			\ar@{.>}@/^.35cm/[drrr]_(.4){\beta_0}|(.29)\hole
			\ar[dr]|(.28)\hole_{y_1\circ e_0} && L_1
			\ar@{.>}@/_.35cm/[dlll]^(.4){\beta_1}
			\ar[dl]|(.28)\hole^{y_1\circ b_0}& K_1 \ar[d]^{d_1\circ
				u_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{g_2\circ
				a_1}\\G'_1 && \ar@{>->}[ll]^{z_1} Q_5 \ar[rr]_{z'_1}&& G'_2 &&
			\ar@{>->}[ll]^{x_2} Q_3 \ar[rr]_{y_2}&& G_3}\]

@@

		Our aim is to construct the dotted arrow in the following diagram.
		\[\xymatrix@C=15pt{L_1 \ar[d]_{f_0\circ i_0}&& K_1
			\ar[d]_{q_0\circ v_1}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_1
			\ar@{.>}@/^.35cm/[drrr]_(.4){\epsilon_0}|(.29)\hole
			\ar[dr]|(.28)\hole_{s_1\circ j_1} && L_2
			\ar@{.>}@/_.35cm/[dlll]^(.4){\epsilon_1}
			\ar[dl]|(.28)\hole^{s_1\circ \alpha_1}& K_2
			\ar[d]^{\lambda_0\circ \phi_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} &&
			R_2 \ar[d]^{\xi_1\circ \gamma_1}\\G_0 && \ar@{>->}[ll]^{s_0} Q_0
			\ar[rr]_{t_0}&& H'_1 && \ar@{>->}[ll]^{\xi_0} Q_6
			\ar[rr]_{\xi'_0}&& H'_2}\]

@@

\begin{proof}
	To fix the notation, let us depict the derivation $\der{D}=\{\dder{D}_i\}_{i=0}^2$:
	\[\xymatrix@C=15pt{L_0\ar[d]^{m_0}&& K_0 \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@/^.35cm/[drrr]_(.4){i_0}|(.29)\hole \ar[dr]|(.28)\hole_{h_0} && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.28)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[r]^{r_1} & R_1 \ar[dr]|(.28)\hole_{h_1} \ar@/^.35cm/[drrr]_(.4){j_0}|(.29)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){j_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]_{h_2} \\G_0 && \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1  && \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2 && \ar@{>->}[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]

@@

	We begin switching $\dder{D}_0$ and $\dder{D}_1$ to get the following diagram.
	\[\xymatrix@C=15pt{L_1\ar[d]^{f_0\circ i_1}&& K_1 \ar[d]_{k'_1}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){a_0}|(.29)\hole \ar[dr]|(.28)\hole_{h'_1} && L_0 \ar@/_.35cm/[dlll]^(.4){a_1} \ar[dl]|(.28)\hole^{m'_0}& K_0 \ar[d]^{k'_0}\ar@{>->}[l]_{l_0} \ar[r]^{r_0} & R_0 \ar[dr]|(.28)\hole_{g_1\circ i_0} \ar@/^.35cm/[drrr]_(.4){b_0}|(.29)\hole  && L_2 \ar@/_.35cm/[dlll]^(.4){b_1} \ar[dl]|(.28)\hole^{m_2}& K_2 \ar[d]^{k_2}\ar@{>->}[l]_{l_2} \ar[rr]^{r_2} && R_2 \ar[d]_{h_2} \\G_0 && \ar@{>->}[ll]^{f'_0} D'_0 \ar[rr]_{g'_0}&& G'_1  && \ar@{>->}[ll]^{f'_1} D'_1 \ar[rr]_{g'_1}&& G_2 && \ar@{>->}[ll]^{f_2} D_2 \ar[rr]_{g_2}&& G_3 }\]
